Implementing Equational Constraints\\ in a Functional Language

### 2011-07-01 papers/wlp2011/paper.tex

@@

\begin{tabular}{|l|c|c|c|}
\hline
Expression: &\code{==} & \code{=:=} & \code{=:<=}\\
\hline
\code{last (take 100000 (repeat failed) ++ [1])} & \code{failed} & \code{failed} & 0.14\\ 
\code{last (map (inc 0) [1..100000])} & stack overflow & 0.45 & 0.13\\
\code{simplify}                       &10.34&10.33&7.27\\
\code{varsInExp} & 0.49& 84.92 & 0.09\\
\code{toPeano 5000 == halfS (toPeano 10000)} & 23.3 & 1.81 & no Fun Pat\\
\hline
\end{tabular}

The table above shows the evaluation of our approach. Primarily we
used the same Benchmarks as in \cite{}

Benchmarks for Last with ==, =:= and func. patterns

also short comparison of =:= with PAKCS and MCC

>>>>

% Redefine description env as equation list to gain key alignment
\usepackage{eqlist}
\let\description=\eqlist
\let\enddescription=\endeqlist
\let\eqlistlabel\descriptionlabel

@@

In Section \ref{sec:Unification} we stated that equational
constraints could also be implemented using pattern matching, but that
this implementation would increase the search space that has to be traversed 
to extract the values of a computation. To show the benificial effects
of our equality constraint implementations, we compare them to
the results using the equality operator \code{==} that
produces such a search space. The programs used for the benchmarks 
are partially taken from \cite{AntoyHanus05LOPSTR}.
%
\begin{quote}
\begin{description}
 \item[last]      Computes the last element of a list
 \item[simplify]  Simplification of symbolic arithmetic expressions
 \item[varInExp]  Nondeterministically returns a variable occuring
                  in a symbolic arithmetic expression
 \item[half]      Computes the half of a peano number using logic variables
 \item[palindrom] Checks if a list is a palindrom
%  \item[grep]
%  \item[horseman]
\end{description}
\end{quote}

Our benchmarks are executed on a Linux machine running Debian 5.0.7 with
an Intel Core 2 Duo (3.0GHz) processor. Our compiler has been executed with
the Glasgow Haskell Compiler (GHC 7.0.4, option -O2). For later comparison with
other systems we considered PAKCS (Version 1.9.2, based on a SICStus-Prolog
4.1.2 backend) and MCC (Version 0.9.10). The timings were performed with the
time command measuring the execution time (in seconds) of a compiled
executable for each benchmark as a mean of three runs.

\begin{figure}
\centering
\begin{tabular}{|l|c|c|c|}
\hline
Expression                              & \code{==} & \code{=:=} & \code{=:<=} \\
\hline
\code{last (map (inc 0) [1..100000])}   & 118.33    & 0.52       & 0.20        \\
% \code{grep}                             & 0.94      & 0.13       &             \\
\code{simplify}                         & 9.18      & 10.33      & 7.09        \\
\code{varInExp}                         & 0.50      & 86.27      & 0.14        \\
\code{fromPeano (half (toPeano 10000))} & 26.49     & 6.26       & 11.58       \\
\code{palindrom}                        & 31.12     & 13.45      & 17.55       \\
% \code{horseman}                         & 3.88      & 4.93       &             \\
\hline
\end{tabular}
\caption{Benchmarks: comparing different types of equation}
 \label{fig:search-strategies}
\end{figure}

\todo{Results from comparing, ==, =:=, =:<=}

As expected, the creation and traversal of a large search space introduced
by \code{==} in most cases is much slower than our presented approach for
unification. In addition, the \code{last} example shows that the lazy 
unification operator \code{=:<=} improves performance when unifying an
expression which has to be evaluted only partially. Using strict unification,
all elements of the list are (unnecessarily) evaluated. Interestingly, in the
\code{varInExp} example the strict unification performs significantly worse
than the other equality operators. \todo{Explain why}

In contrast to the Curry implementations PAKCS and MCC, out approach is based
upon an explicit representation of the search space and does not manipulate
a global state for unifying logic variables.
Nevertheless, our benchmarks show that it can in most cases
compete or even outperform these implementations.
Figure~\ref{fig:unification} shows a comparison of the curry systems where
strict equality (\ccode{=:=}) is used for unification.\footnote{%
\ccode{inc x n} is a naive addition that \code{n} times
increases its argument \code{x} by 1.} We can deduce from the results
that the unification of the MCC performs best. However, we can in most cases
outperform the Prolog based PAKCS but for some examples where further
research and optimization is necessary.

\begin{figure}
\centering
\begin{tabular}{|l|c|c|c|}
\hline
Expression                              & KiCS2 & PAKCS & MCC  \\
\hline
\code{last (map (inc 0) [1..100000])}   & 0.52  & 4.27   & 0.15 \\
% \code{grep}                             & 0.13  & 2.83   & 0.15 \\
\code{simplify}                         & 10.33 & 0.15   & 0.01 \\
\code{varInExp}                         & 86.27 & 0.35   & 0.03 \\
\code{fromPeano (half (toPeano 10000))} & 6.26  & 109.71 & 3.17 \\
\code{palindrom}                        & 13.45 & 32.82  & 1.04 \\
% \code{horseman}                         & 4.93  & 8.85   & 0.42 \\
\hline
\end{tabular}
\caption{Benchmarks: strict equational constraints}
 \label{fig:unification}
\end{figure}

As the MCC does not support functional patterns, the performance of lazy
unification is compared with PAKCS only (Fig.~\ref{fig:functional-patterns}).
Again, our compiler performs well against PAKCS
and outperforms it in most cases. \todo{extend}

\begin{figure}
\centering
\begin{tabular}{|l|c|c|}
\hline
Expression                              & KiCS2 & PAKCS    \\
\hline
\code{last (map (inc 0) [1..100000])}   & 0.20  & 3.93     \\
\code{simplify}                         & 7.09  & 0.26     \\
\code{varInExp}                         & 0.14  & 0.69     \\
\code{fromPeano (half (toPeano 10000))} & 11.57 & $\infty$ \\
\code{palindrom}                        & 17.55 & $\infty$ \\
\hline
\end{tabular}
\caption{Benchmarks: functional patterns}
 \label{fig:functional-patterns}
\end{figure}
